Exploring the tree of numerical semigroups

### 2014-09-05 Papier/article.tex
Biblio + new attempt of curves.

\noindent In this version we can see that the cursors \texttt{src} and
\texttt{dst} move at the same time and that the modification of the value
pointed by \texttt{dst} only needs to access the values pointed by \tt{src}
and \tt{dst}.  We can therefore work in multiple entries at the same time
without collision.  Current CPUs allow this thanks to the \SIMD technologies as
\MMX, \SSE, etc.  The acronym \SIMD~\cite{SIMD} stands for Single Operation
Multiple Data.  We used \SSEV~\cite{SSE, IntelSSE} technology as it allows
for the largest speedup. This need to respect some constraints in the memory
organization of the data, namely ``memory alignment''. Recall that an address
is 16~bytes aligned if it is a multiple of 16. \SSE memory access are much
faster for aligned access.

@@

As we previously said, to gain more speed this core loop is actually unrolled
using some kind of Duff device~\cite{Duff} as
\begin{algorithmic}
\State \textbf{Jump to case } $\tt{start}+1$
\State \textbf{case 1:} \tt{block = ...; dst[1] -= ((block != zero) \& one)}
\State \textbf{case 2:} \tt{block = ...; dst[2] -= ((block != zero) \& one)}
\State \textbf{[...]}
\State \textbf{case NBLOCK-1:} \tt{...}
\end{algorithmic}

@@

Our second optimization is to use parallelism on exploration of the tree.
Today, CPU of personal computer have several cores (2, 4 or more).  The given
version of our exploration algorithm uses a single core and so a fraction only
of the power of a CPU. The idea here is that different branches of the tree can
be explored in parallel by different cores of the computer. The tricky part is
to ensure that each cores are busy, giving a new branch when a core is done
with a former one. Fortunately there is a technology called
\CilkP~\cite{CilkIntel, CilkRefman}
which is particularly well suited for those kind of problems.

@@

The thing are actually only a little bit more complicated. First we have to
gather the results of the exploration. If we simply write
\begin{algorithmic}[1]
\State \tt{result}[$g(\tt{S})$] $\gets \tt{result}[g(\tt{S})] + 1$
\end{algorithmic}
then we face the problem of two cores incrementing the same variable at the
same time.
 Incrementing a variable is actually done in 3 steps: reading the
value from the memory, adding one, storing back the result. 
Since there is by
default non synchronization, the following sequence of actions for two cores
 is possible: Read1 / Read2 / Add1 / Add2 / Store1 / Store2. Then the two
cores performs the same modification resulting in incrementing the variable
only once. This is called a data race and leads to nondeterministic wrong
result. To cope with those kinds of synchronization problems, \Cilk provide the
notion of \emph{reducer} which are variables local to each thread which are
gathered (here added) when a thread is finishing its job.

@@

\begin{figure}[b]
\begin{tikzpicture}[scale=1.3]
        \begin{semilogyaxis}[xlabel=\small genus,ylabel=time,legend pos=north west,legend cell align=left]
        \addplot [color=blue,mark=x] coordinates {
            (20,0.02)
            (21,0.04)
            (22,0.07)
            (23,0.12)
            (24,0.2)
            (25,0.35)
            (26,0.61)
            (27,1.04)
            (28,1.76)
            (29,2.96)
            (30,4.96)
            (31,8.26)
            (32,13.77)
            (33,22.95)
            (34,38.22)
            (35,1250.88)
        };
       \addplot [color=black,mark=x] coordinates {
			(20,0.02)	
            (21,0.04)
            (22,0.06)
            (23,0.09)
            (24,0.16)
            (25,0.26)
            (26,0.45)
            (27,0.72)
            (28,1.25)
            (29,2.1)
            (30,3.39)
            (31,5.85)
            (32,9.25)
            (33,16.34)
            (34,27.23)
            (35,45.39)
            (36,75.58)
            (37,125.59)
            (38,203.56)
            (39,346.02)
            (40,557.51)
        };
         \addplot [color=red,mark=x] coordinates {
			(20,0.02)	
            (21,0.02)
            (22,0.04)
            (23,0.06)
            (24,0.1)
            (25,0.18)
            (26,0.29)
            (27,0.48)
            (28,0.81)
            (29,1.34)
            (30,2.25)
            (31,3.73)
            (32,6.2)
            (33,10.35)
            (34,17.11)
            (35,28.35)
            (36,46.99)
            (37,77.89)
            (38,128.53)
            (39,212.1)
            (40,349.98)
        };
        \addplot [color=magenta,mark=x] coordinates {
			(20,0.01)	
            (21,0.01)
            (22,0.01)
            (23,0.01)
            (24,0.02)
            (25,0.03)
            (26,0.05)
            (27,0.08)
            (28,0.12)
            (29,0.21)
            (30,0.35)
            (31,0.57)
            (32,0.97)
            (33,1.65)
            (34,2.69)
            (35,4.18)
            (36,7.36)
            (37,11.66)
            (38,20.35)
            (39,32.55)
            (40,74.46)
        };
	%\legend style={anchor=north,at={(20,1000)}}  
        \legend{\tiny breadth-expl,\tiny depth-expl,\tiny depth-expl-$\delta$,\tiny depth-expl-$\delta$-mmx}
        \end{semilogyaxis}
    \end{tikzpicture}
    \label{F:Time}
    \caption{Comparison of time executions of different exploration algorithms of the tree $\mathcal{T}^{\leq g}$ on an Intel \tt{i5-3570K} CPU with $8GB$ of memory. 
The scale is logarithmic in time. 
The first version is based on a breadth-first search exploration and the peak at genus~$35$ is due to the consumption of all the memory and use of swap. 
The second version uses a depth first search exploration.
The third is based on the second one with use of decomposition numbers, it corresponds to the algorithm given in Section~\ref{S:Algo}. 
The last one is an optimisation of the third one with \MMX.}
\end{figure}

>>>>

\noindent In this version we can see that the cursors \texttt{src} and
\texttt{dst} move at the same time and that the modification of the value
pointed by \texttt{dst} only needs to access the values pointed by \tt{src}
and \tt{dst}.  We can therefore work in multiple entries at the same time
without collision.  Current CPUs allow this thanks to the \SIMD technologies as
\MMX, \SSE, etc.  The acronym \SIMD~\cite{WikipediaSIMD} stands for Single Operation
Multiple Data.  We used \SSEV~\cite{WikipediaSSE, IntelSSE} technology as it allows
for the largest speedup. This need to respect some constraints in the memory
organization of the data, namely ``memory alignment''. Recall that an address
is 16~bytes aligned if it is a multiple of 16. \SSE memory access are much
faster for aligned access.

@@

As we previously said, to gain more speed this core loop is actually unrolled
using some kind of Duff device~\cite{WikipediaDuff} as
\begin{algorithmic}
\State \textbf{Jump to case } $\tt{start}+1$
\State \textbf{case 1:} \tt{block = ...; dst[1] -= ((block != zero) \& one)}
\State \textbf{case 2:} \tt{block = ...; dst[2] -= ((block != zero) \& one)}
\State \textbf{[...]}
\State \textbf{case NBLOCK-1:} \tt{...}
\end{algorithmic}

@@

Our second optimization is to use parallelism on exploration of the tree.
Today, CPU of personal computer have several cores (2, 4 or more).  The given
version of our exploration algorithm uses a single core and so a fraction only
of the power of a CPU. The idea here is that different branches of the tree
can be explored in parallel by different cores of the computer. The tricky
part is to ensure that each cores are busy, giving a new branch when a core is
done with a former one. Fortunately there is a technology called
\CilkP~\cite{CilkIntel, CilkRefman} which is particularly well suited for
those kind of problems. For our computation, we used the free version which is
integrated in the latest version of the GNU C compiler~\cite{GCCcilk}.

@@

The thing are actually only a little bit more complicated. First we have to
gather the results of the exploration. If we simply write
\begin{algorithmic}[1]
\State \tt{result}[$g(\tt{S})$] $\gets \tt{result}[g(\tt{S})] + 1$
\end{algorithmic}
then we face the problem of two cores incrementing the same variable at the
same time.
 Incrementing a variable is actually done in 3 steps: reading the
value from the memory, adding one, storing back the result. 
Since there is by
default non synchronization, the following sequence of actions for two cores
 is possible: Read1 / Read2 / Add1 / Add2 / Store1 / Store2. Then the two
cores performs the same modification resulting in incrementing the variable
only once. This is called a data race and leads to nondeterministic wrong
result. To cope with those kinds of synchronization problems, \CilkP provide the
notion of \emph{reducer} which are variables local to each thread which are
gathered (here added) when a thread is finishing its job.

@@

\begin{figure}[b]
\begin{tikzpicture}[scale=1.3]
        \begin{semilogyaxis}[xlabel=\small genus,ylabel=time,legend pos=north west,legend cell align=left]
        \addplot [color=blue,mark=x] coordinates {
            (20,0.02)
            (21,0.04)
            (22,0.07)
            (23,0.12)
            (24,0.2)
            (25,0.35)
            (26,0.61)
            (27,1.04)
            (28,1.76)
            (29,2.96)
            (30,4.96)
            (31,8.26)
            (32,13.77)
            (33,22.95)
            (34,38.22)
            (35,1250.88)
        };
       \addplot [color=black,mark=x] coordinates {
			(20,0.02)	
            (21,0.04)
            (22,0.06)
            (23,0.09)
            (24,0.16)
            (25,0.26)
            (26,0.45)
            (27,0.72)
            (28,1.25)
            (29,2.1)
            (30,3.39)
            (31,5.85)
            (32,9.25)
            (33,16.34)
            (34,27.23)
            (35,45.39)
            (36,75.58)
            (37,125.59)
            (38,203.56)
            (39,346.02)
            (40,557.51)
        };
         \addplot [color=red,mark=x] coordinates {
			(20,0.02)	
            (21,0.02)
            (22,0.04)
            (23,0.06)
            (24,0.1)
            (25,0.18)
            (26,0.29)
            (27,0.48)
            (28,0.81)
            (29,1.34)
            (30,2.25)
            (31,3.73)
            (32,6.2)
            (33,10.35)
            (34,17.11)
            (35,28.35)
            (36,46.99)
            (37,77.89)
            (38,128.53)
            (39,212.1)
            (40,349.98)
        };
        \addplot [color=magenta,mark=x] coordinates {
			(20,0.01)	
            (21,0.01)
            (22,0.01)
            (23,0.01)
            (24,0.02)
            (25,0.03)
            (26,0.05)
            (27,0.08)
            (28,0.12)
            (29,0.21)
            (30,0.35)
            (31,0.57)
            (32,0.97)
            (33,1.65)
            (34,2.69)
            (35,4.18)
            (36,7.36)
            (37,11.66)
            (38,20.35)
            (39,32.55)
            (40,74.46)
        };
	%\legend style={anchor=north,at={(20,1000)}}  
        \legend{\tiny breadth-expl,\tiny depth-expl,\tiny depth-expl-$\delta$,\tiny depth-expl-$\delta$-mmx}
        \end{semilogyaxis}
    \end{tikzpicture}
    \label{F:Time:new}
    \caption{Comparison of time executions of different exploration algorithms of the tree $\mathcal{T}^{\leq g}$ on an Intel \tt{i5-3570K} CPU with $8GB$ of memory. 
The scale is logarithmic in time. 
The first version is based on a breadth-first search exploration and the peak at genus~$35$ is due to the consumption of all the memory and use of swap. 
The second version uses a depth first search exploration.
The third is based on the second one with use of decomposition numbers, it corresponds to the algorithm given in Section~\ref{S:Algo}. 
The last one is an optimisation of the third one with \MMX.}
\end{figure}

\begin{figure}[b]
\begin{tikzpicture}[scale=1.3]
        \begin{semilogyaxis}[xlabel=\small genus,ylabel=time,legend pos=north west,legend cell align=left]
        \addplot [color=blue,mark=x] coordinates {
(20,0.045767)
(21,0.067697)
(22,0.112697)
(23,0.194646)
(24,0.335412)
(25,0.581481)
(26,1.01088)
(27,1.72613)
(28,2.9324)
(29,4.91856)
(30,8.39383)
(31,14.0308)
(32,23.7138)
(33,39.6461)
(34,66.1154)
(35,109.406)
(36,182.331)
(37,302.765)
        };
       \addplot [color=black,mark=x] coordinates {
			(20,0.02)	
(20,0.043697)
(21,0.065487)
(22,0.113301)
(23,0.186903)
(24,0.31748)
(25,0.53251)
(26,0.899982)
(27,1.5205)
(28,2.55423)
(29,4.28809)
(30,7.30236)
(31,12.1284)
(32,20.2132)
(33,33.707)
(34,56.5415)
(35,93.7614)
(36,157.09)
(37,259.997)
(38,427.497)
(39,709.564)
(40,1192.81)
        };
         \addplot [color=red,mark=x] coordinates {
(20,0.027708)
(21,0.044416)
(22,0.074789)
(23,0.122006)
(24,0.203155)
(25,0.344143)
(26,0.594073)
(27,1.07134)
(28,1.622)
(29,2.82573)
(30,4.53882)
(31,7.60975)
(32,12.5442)
(33,21.0466)
(34,34.7638)
(35,57.7621)
(36,95.6878)
(37,158.95)
(38,264.495)
(39,436.146)
(40,719.347)
        };
        \addplot [color=magenta,mark=x] coordinates {

        };
	%\legend style={anchor=north,at={(20,1000)}}  
        \legend{\tiny breadth-expl,\tiny depth-expl,\tiny depth-expl-$\delta$}
        \end{semilogyaxis}
    \end{tikzpicture}
    \label{F:Time}
    \caption{Comparison of time executions of different exploration algorithms of the tree $\mathcal{T}^{\leq g}$ on an Intel \tt{i5-3570K} CPU with $8GB$ of memory. 
The scale is logarithmic in time. 
The first version is based on a breadth-first search exploration and the peak at genus~$35$ is due to the consumption of all the memory and use of swap. 
The second version uses a depth first search exploration.
The third is based on the second one with use of decomposition numbers, it corresponds to the algorithm given in Section~\ref{S:Algo}. 
The last one is an optimisation of the third one with \MMX.}
\end{figure}
